---
  title: "Vocabulary for Chapter 4"
author: "Amy Fox"
date: '2020-01-13'
slug: vocabulary-for-chapter-4
categories:
  - vocabulary
- Chapter 4
tags:
  - vocabulary
- Chapter 4
subtitle: ''
summary: ''
authors: []
lastmod: 
  featured: no
image:
  caption: ''
focal_point: ''
preview_only: no
projects: []
---
  
Chapter 4 covers distributions and how to estimate parameters that define distributions. It introduces a number of terms covering distributions, as well as a few biological terms. The vocabulary words for 
Chapter 4 are: 
  
```{r echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(readr)
vocab <- read_tsv("vocab_lists/chapter_4.tsv", col_names = c("term", "definition"))
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(kableExtra)
vocab %>% 
  kable(align = c("rl"), 
        col.names = c("", "")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover",
                                      "condensed")) %>% 
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2, width = "30em")
```

### Sources consulted or cited 

*Some of the definitions above are based in part or whole on listed definitions 
in the following sources.*
  
  - Holmes and Huber, 2019. *Modern Statistics for Modern Biology.* Cambridge University Press, 
Cambridge, United Kingdom.
- Everitt and Skrondal, 2010. *The Cambridge Dictionary of Statistics (Fourth Edition).* Cambridge University Press, Cambridge, United Kingdom.
- *Zero-Inflated Poisson Regression.* Institute for Digital Research and Education Statistical Consulting. stats.idre.ucla.edu.
- Berrar, 2019. *Introduction to Non-parametric Bootstrap.* Research Gate.


### Practice

<iframe src="https://quizlet.com/473473135/flashcards/embed?i=2ldef3&x=1jj1" height="500" width="100%" style="border:0"></iframe>