---
title: Chapter 1 exercise setup
author: Bailey Fosdick
date: '2020-02-06'
slug: chapter-1-exercise-setup
categories:
  - Chapter 1
  - exercises
tags:
  - Chapter 1
subtitle: 'Exercise 1.8 setup'
summary: 'Instructions on how to get started on Chapter 1, exercise 1.8.'
authors: []
lastmod: '2020-02-06T10:05:38-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

The code instructions in the exercise statement appear to be outdated.  The code below worked on my machine.  Note that when asked whether I would like to update packages from the binary version, I said no.  (When I said yes, `R` gave an error.)

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("Biostrings", "BSgenome.Celegans.UCSC.ce2","BSgenome"))
```

You can see the various data genome data sets available by loading the `BSgenome` library and typing `available.genomes()`. 


Once you have the needed packages installed, you can access the sequence data for this exercise via the following commands.
```{r}
suppressMessages(library("BSgenome.Celegans.UCSC.ce2"))
Celegans
seqnames(Celegans)
Celegans$chrM
class(Celegans$chrM)
length(Celegans$chrM)
```

The Biostrings packages provides functions to summarize the sequence.  For example:
```{r}
library("Biostrings")
lfM = letterFrequency(Celegans$chrM, letters=c("A", "C", "G", "T"))
lfM
sum(lfM)
```
