---
title: "Exercise Solution for Chapter 6"
author: "Sherry WeMott"
date: '2020-05-12'
output: pdf_document
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
lastmod: '2020-05-12T10:46:22-06:00'
categories:
- Chapter 6
- exercises
projects: []
slug: exercise-solution-for-chapter-6
subtitle: ''
summary: ''
tags:
- Chapter 6
- exercises
authors: []
---

## Chapter 6, Exercise 6.4

We are instructed to make a less extreme example of correlated test statistics than the data duplication at the end of Section 6.5. Simulate data with true null hypotheses only, and let the data morph from having completely independent replicates (columns) to highly correlated as a function of some continuous-valued control parameter. Check type-I error control (e.g., with the p-value histogram) as a function of this control parameter.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=F, message=F, warning=F}
library(tidyverse)
library(purrr)
library(ggbeeswarm)
```

For this exercise we use the PlantGrowth dataset from the `datasets` package in R. The dataset includes results from an experiment on plant growth comparing yeilds (as measured by dried weight of plants) obtained under a control and two different treatment conditions. 

```{r}
data("PlantGrowth")
PlantGrowth
```

Here we're plotting the outcomes of the three groups (ctrl, trt1, and trt2)

```{r fig.height = 2, fig.width = 3, fig.align = "center"}
PlantGrowth %>% 
   mutate(group = fct_recode(group, 
                            Control = "ctrl", 
                            `Treatment 1` = "trt1", 
                            `Treatment 2` = "trt2")) %>% 
  ggplot(aes(x = group, y = weight)) + 
  geom_beeswarm() + 
  labs(x = "")
```

Next we'll apply a t-test comparing weights in the ctrl group to the trt2 group: 

```{r}
PlantGrowth %>% 
  filter(group %in% c("ctrl", "trt2")) %>% 
    t.test(weight ~ group, data  = .) 
```
Here's the tidy version: 

```{r}
library("broom")
PlantGrowth %>% 
  filter(group %in% c("ctrl", "trt2")) %>% 
  t.test(weight ~ group, data  = .) %>% 
    tidy()
```

Then we'll duplicate data, adding a second copy of the dataframe, before running the t-test. Notice taht the p-value is smaller even though the group means haven't changed. This is due to the increase in sample size. 

```{r}
PlantGrowth %>% 
  bind_rows(PlantGrowth) %>% # Add the duplicate of the dataset
  filter(group %in% c("ctrl", "trt2")) %>% 
  t.test(weight ~ group, data  = .) %>% 
  tidy()
```

Here we resample only half the data: 

```{r}
PlantGrowth %>% 
  sample_frac(size = 0.5) %>% 
  bind_rows(., .) %>% 
  filter(group %in% c("ctrl", "trt2")) %>% 
  t.test(weight ~ group, data  = .) %>% 
  tidy()
```

Add random noise: 

```{r}
pg1 <- PlantGrowth %>% 
  sample_frac(size = 0.5) 
pg2 <- pg1 %>% 
  mutate(noise = rnorm(15, mean = 0, sd = 0.2), 
         weight = weight + noise) %>% 
  select(-noise)

pg1 %>% 
  bind_rows(pg2) %>% 
  filter(group %in% c("ctrl", "trt2")) %>% 
  t.test(weight ~ group, data  = .) %>% 
  tidy()
```
```{r}

```
