---
title: "Exercise Solution for Chapter 12"
author: "Amy Fox"
date: '2020-05-12'
slug: exercise-solution-for-chapter-12
draft: no
categories:
  - Exercise solutions
  - Chapter 12
tags:
  - Exercise solutions
  - Chapter 12
subtitle: ''
summary: ''
authors: [amy-fox]
lastmod:
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 12.2 from Modern Statistics for Modern Biologists

> Use glmnet for a prediction of a continous variable, i.e., for regression. Use the prostate cancer data from Chapter 3 of (Hastie, Tibshirani, and Friedman 2008). The data are available in the CRAN package `ElemStatLearn`. Explore the effects of using Ridge versus Lasso penalty.


Here are the packages that need to be installed.
```{r message = FALSE}
library(dplyr)
library(ggplot2)
library(glmnet) # perform generalize linear models
library(GGally) # used for ggpairs function
library(superheat) # used to show correlation between variables
```

## Data for the exercise

The `ElemStatPackage` isn't on CRAN anymore. But it is possible to download it from Github at: https://github.com/cran/ElemStatLearn/blob/master/data/prostate.RData

```{r}
load("./example_datasets/prostate.RData")

prostate %>% 
  head()
```

This data looks at correlating the level of prostate-specific antigen `lpsa` and eight other clinical measures in men.

Here's what the variables mean: 

- `lcavol`: log cancer volume
- `lweight`: log prostate weight
- `age`: in years
- `lbph`: log of the amount of benign prostatic hyperplasia
- `svi`: seminal vesicle invasion
- `lcp`: log of capsular penetration
- `gleason`: a numeric vector with the Gleason score
- `pgg45`: percent of Gleason score 4 or 5
- `lpsa`: response (the thing you are trying to predict), the 
level of prostate-specific antigen
- `train`: a logical vector, of whether the data was to be 
part of the training dataset (TRUE) or the testing one (FALSE)

So, you're trying to predict the values of `lpsa` based on all of the other variables. `lpsa` is a continuous variable. To get a general view of the `lpsa` values, we can view a summary of values and create a histogram.

```{r}
prostate$lpsa %>%
  summary()


ggplot(prostate, aes(x = lpsa)) +
  geom_histogram(bins = 30) +
  ggtitle("Histogram of lpsa response")
```

Here we can see the values range from -0.4 to values of almost 6. Based on the histogram, the data seems to also fit a close to normal distribution.

Before actually performing any predictions, we may want to see if any of the variables are correlated to each other. This can show if there are potentially two variables measuring similar things. We use the `ggpairs` function from the `GGally` package to do this.

```{r message = FALSE, fig.width = 6, fig.height = 8}
ggpairs(prostate)
```

Based on this plot, we can see that some variables, such as`pgg45` and `gleason`, are highly correlated. We can also see the distribution of the data in the testing and training datasets (we will discuss the testing and training column later in this exercise).

We can also visually view correlations between the variables using the `cor` function and a heatmap from the`superheat` package.

```{r}
prostate %>%
  select(-train) %>%
  cor() %>%
  superheat()
```

This plot shows the same data as the `ggpairs` output, but we can visually see the correlation between each of the variables. If the correlation is high and positive, then the rectange is yellow, if the correlation between two variables is low, the rectangle is purple.

We will first split the data into testing and training data.
```{r}
prostate_train <- prostate %>%
  filter(train == TRUE)

prostate_test <- prostate %>%
  filter(train == FALSE)
```

There are `r nrow(prostate_train)` samples in the training set and `r nrow(prostate_test)` samples in the testing set.


Because this data was specifically posted to teach machine learning, the authors included a column called `train` so that users can split the data into testing and training datasets. If this column had not been created, there are many ways to sample the data and split it into two datasets. Here is an example:

`install.packages("rsample")`
`library(rsample)`
`prostate_split <- initial_split(prostate, prop = .75)`
`prostate_test <- testing(prostate_split)`
`prostate_train <- training(prostate_split)`



## Fit generalized linear model (glmnet) with Lasso and Ridge penalties

We will first pull out the predictors and reponse that we want to input into the `glmnet` function.

```{r}
# pull outall of the predictors that we are using
predictors <- prostate_train %>%
  select(lcavol:pgg45) %>%
  as.matrix()

head(predictors)

# this is the response
response <- prostate_train %>%
  pull(lpsa)
```

When performing statistical analysis on data with many variables, we often have a problem with variance as there are many parameters. We can use penalization to account for this variance-bias trade off. Two examples of penalization are the Lasso and Ridge penalty which we will use in this exercise.

Based on the `glmnet` package, when `alpha = 1` the Lasso penalty is used, if `alpha = 0`, then Ridge penalty is used. A great resource for the `glmnet` package can be found here: <https://web.stanford.edu/~hastie/glmnet/glmnet_alpha.html>

Here, we use a matrix of all of the predictors (`x`) to try to predict the `lpsa` column (`y`). When we plot the penalized regressions, we use the `label = TRUE` to show which lines correspond to which predictors. In this case, the order of the input vector `predictors` matches with the variable numbers as follows. 

1: lcavol

2: lweight

3: age

4: lbph

5: svi

6: lcp

7: gleason

8: pgg45

```{r}

# Lasso 
lasso_glmnet <- glmnet(x = predictors, 
                    y = response, 
                    family = "gaussian", alpha = 1)

plot(lasso_glmnet, label = TRUE)
title("Lasso Penalty", line = -2.5)

# Ridge
ridge_glmnet <- glmnet(x = predictors, 
                    y = response, 
                    family = "gaussian", alpha = 0)

plot(ridge_glmnet, label = TRUE)
title("Ridge Penalty", line = -1.5)
```

The plots show the estimated coefficients as L1 norm increases. L1 norm is the regularization term. It means that at small L1 norm values, there is a lot of regularization, but as it increases, more variables become part of the model as coefficients take non-zero values. The top axis shows the number of nonzero coefficients correspoding to the lamba at the current L1 norm. From these plots we can see that the variables 1 (log cancer volume), 2 (log prostate weight), and 5 (seminal vesicle invasion) are good predictors for the level of prostate-specific antigen (lpsa response).


## Cross Validation

After creating our models, we want to see how good they are. We can do this through cross-validation. In this case, we already split the data into training and testing data, but we could run the full modeling and cross-validation on all of the data. We use the `cv.glmnet` function to do this. It utilizes k-fold cross validation, meaning that the input data is split multiple times into new trainng and testing sets of sizes n(k-1)/k and n/k, respectively. We use this cross-validation to determine the best $\lambda$. Again we input a matrix of all of the predictors (`x`) to look at the `lpsa` response (`y`). 

```{r}
set.seed(2)

# Lasso
cvglmnet_lasso <- cv.glmnet(x = predictors, 
                    y = response, 
                    family = "gaussian", alpha = 0)
cvglmnet_lasso

plot(cvglmnet_lasso)
title("Lasso Cross Validation", line = -1.5)

# Ridge
cvglmnet_ridge <- cv.glmnet(x = predictors, 
                    y = response, 
                    family = "gaussian", alpha = 1)
cvglmnet_ridge

plot(cvglmnet_ridge)
title("Ridge Cross Validation", line = -1.5)
```

In the data output, `1se` is the data point that is within 1 standard error of the minimum $\lambda$ (`min`). This is the value that the model suggests that we use in our ultiamate predictive model (indicated by the 2nd dotted line on the plots.)
The `1se Measure`  is similar to the mean squared error. If the measure is small, then the model is better. When comparing the `Measure` of the `1se` between the two penalties, we can see that the Ridge Penalty has a smaller `1se Measure`, showing that it performs better. 

The `Nonzero` column describes the nonzero coefficients, or the number of predictors that are important in the particular model. There were a total of 8 predictors as the input. The Lasso penalty shows that all 8 predictors are important in building the model, but the Ridge penalty only uses 5 predictors.

Based on this cross-validation, we can deduce that the Ridge penalty is the best predictive model to use for this data. We have also identified the best $\lambda$ to use with this model at 0.18074.

## Lasso Prediction 

Here we pull out the same variables that we used to create the model, but we use the values from the testing data instead of the training data. We will then use the predictors from the testing data to predict the `lpsa`.
```{r}
predictors_test <- prostate_test %>% 
  dplyr::select(lcavol:pgg45) %>% 
  as.matrix()

head(predictors_test)
```


As we used the training data to build the model, we can then test the generalized linear model with the Lasso penalty on the testing data.

We start by using the `predict` function to use the model to predict the `lpsa` on the testing data. We can then see the correlation between the predicted values and actual values.
```{r}
s0 <- cvglmnet_lasso$lambda.1se 

lasso_predict <- predict(cvglmnet_lasso, newx = predictors_test, s = s0)

# create a data frame of the actual lpsa values and the predicted lpsa values
actual_lasso_predict_df <- data.frame(prediction = as.vector(lasso_predict), actual = prostate_test$lpsa)

```

We can then see how correlated the prediction and real data are using the `cor` function.
```{r}
# look at the correlation of the prediction and real data
cor(actual_lasso_predict_df)
```
The output shows that the actual and predicted values are 72% correlated.

Finally, we can plot the actual vs. predicted values on a scatter plot. If the actual and predicted values match up exactly, they would sit on the y = x line.
```{r}
ggplot(actual_lasso_predict_df, aes(x = actual, y = prediction)) +
  geom_point(color = "#00B0F6", size = 2) +
  geom_abline(slope=1, intercept=0)+
  ggtitle("Lasso Prediction") +
  theme_light()+
  coord_fixed(xlim = c(0.75, 5.6),
              ylim = c(0.75, 5.6))
```

We can see in regions of the plot above the y = x line that the model overpredicted the actual values. If a point is below the y = x line, the model underpredicted the values.

## Ridge Prediction 

We can then perform the same functions using the  generalized linear model with the Ridge penalty to test on the testing data.

```{r}
s0 <- cvglmnet_ridge$lambda.1se 

ridge_predict <- predict(cvglmnet_ridge, newx = predictors_test, s = s0)

# create a data frame of the predicted values and actual values
actual_ridge_predict_df <- data.frame(prediction = as.vector(ridge_predict), actual = prostate_test$lpsa) 
```

We can then see how correlated the prediction and real data are using the `cor` function again.
```{r}
cor(actual_ridge_predict_df)
```
The Ridge prediction and actual values are 77% correlated.

Finally, we can plot the actual vs. predicted values on a scatter plot. If the actual and predicted values matched up exactly, they would sit on the y = x line.
```{r}
ggplot(actual_ridge_predict_df, aes(x = actual, y = prediction)) +
  geom_point(color = "#FF62BC", size = 2) +
  geom_abline(slope=1, intercept=0) +
  ggtitle("Ridge Prediction") +
  theme_light() +
  coord_fixed(xlim = c(0.75, 5.6),
              ylim = c(0.75, 5.6))
```

We can see in regions of the plot above the y = x line that the model overpredicted the actual values. If a point is below the y = x line, the model underpredicted the values.

## Conclusion

Comparing the Lasso and Ridge penalty, based on the cross-validation, the Ridge penalty had a smaller `1se Measure`, showing that it performs better. When looking at the actual predictions on the testing data, the Ridge penalty had a higher correlation between the predicted and actual values (77%) compared to the Lasso penalty correlation (72%). In conclusion, the **Ridge penalty performed better on this particular data set.**


Resources
- https://stats.stackexchange.com/questions/68431/interpretting-lasso-variable-trace-plots