---
title: Exercise solution for Chapter 1
author: Brooke Anderson
date: '2020-02-11'
slug: exercise-solution-for-chapter-1
categories:
  - exercises
  - Chapter 1
tags:
  - exercises
  - Chapter 1
subtitle: ''
summary: ''
authors: [brooke-anderson]
lastmod: '2020-02-11T20:22:07-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---


This exercise asks us to explore the frequency of each of the four nucleotides
(A, C, G, and T) in the genome of *C. elegans*, a type of worm used frequently
in scientific research.

This solution requires that several R extension packages be loaded in your R
session. If you do not have these packages installed to your computer yet, you
should follow [instructions we've posted
separately](https://kind-neumann-789611.netlify.com/post/chapter-1-exercise-setup/)
describing the required set-up for this exercise. Once you have installed these
packages on your computer, you can load them into your current R session using
the `library` function:

```{r message = FALSE}
library("BSgenome.Celegans.UCSC.ce2")
library("Biostrings")

library("tidyverse")
library("knitr")
```

## Part A

Part A of the question asks us to explore the nucleotide frequency of the *C.
elegans* genome. This genome is available in the `Celegans` data that comes with
the `BSgenome.Clegans.UCSC.ce2` package and is stored within a `BSgenome` class,
which is a special object class provided by the `Biostrings` package.

There is a dedicated function called `letterFrequency` in the `Biostrings`
package that can be used to count the frequency of letters in a string (like a
genome) in an R object like this. In a call to this function, you must also
include the possible letters in your "alphabet"---that is, the possible letters
that each position in your string could take.

```{r}
(nuc_freq <- letterFrequency(Celegans$chrM, letters=c("A", "C", "G", "T")))
```

To explore and plot this data, I put this summary data into a `tibble`, so I
could more easily use `tidyverse` tools with the data.

```{r}
nuc_freq_df <- tibble(nucleotide = names(nuc_freq), 
             n = nuc_freq)
nuc_freq_df
```

In this format, you can use `tidyverse` tools to explore the data a bit more.
For example, you can determine the total number of nucleotides in the genome
and, with that calculate the proportion of each nucleotide across the genome.
Along with the `kable` function from the `knitr` package, I created a formatted
table with this information:

```{r}
nuc_freq_df %>% 
  mutate(prop = n / sum(n)) %>% 
  kable(digits = 2, 
        caption = "Nucleotide frequencies and proportions in *C. elegans*",
        col.names = c("Nucleotide", "Frequency", "Proportion"))
```

For some presentations, it might be clearer to present this information in a
slightly different table format, using `pivot_longer` and then `pivot_wider` to
reformat the table for presentation:

```{r}
nuc_freq_df %>% 
  mutate(prop = n / sum(n),
         n = prettyNum(n, big.mark = ","),
         prop = prettyNum(prop, digits = 2)) %>% 
  pivot_longer(cols = c("n", "prop")) %>% 
  pivot_wider(names_from = "nucleotide") %>% 
  mutate(name = case_when(
    name == "n" ~ "Frequency of nucleotide",
    name == "prop" ~ "Proportion of all nucleotides"
  )) %>%  
  rename(` ` = name) %>% 
  kable(align = c("rcccc"), 
        caption = "Nucleotide frequencies and proportions in *C. elegans*")
```

Here is a plot of the frequency of each of the four nucleotides for the *C.
elegans* nucleotide:

```{r}
ggplot(nuc_freq_df, aes(x = nucleotide, y = n)) + 
  geom_col(fill = "lavender", color = "black") + 
  theme_classic() + 
  scale_y_continuous(label = scales::comma) + 
  theme(axis.title = element_blank()) + 
  labs(title = expression(paste(italic("C. elegans"), " neucleotide frequency")),
       caption = expression(paste("Based on data from the ", italic("BSgenome.Celegans.UCSC.ce2"), 
                                  " package.")))
```

This graph uses a few elements to improve its appearance that you might want to
explore if you're not already familiar with them:

- The `labs` function is used to add both a title and a caption to the plot.
- The `paste`, `expression`, and `italic` functions are used together to put "C.
elegans" and an R package name in italics in some of the labels on the plot.
- The `scales` package is used inside a scale layer for the `ggplot2` code to
make the y-axis labels a bit nicer.
- `theme` calls are used to apply a simpler overall theme than the default and to 
remove the x- and y-axis titles (with `element_blank`).
- The color and fill of the bars are customized in the geom layer (`geom_col`).

From this plot, it certainly looks like the nucleotides are **not** uniformly
distributed in the *C. elegans* genome. This question will be investigated more
in the next part of the exercise.

## Part B

The second part of the exercise asks us to test whether the observed nucleotide
data for *C. elegans* is  consistent with the uniform model that all nucleotide
frequencies are the same.

First, we can simulate several datasets under this null model and see how a plot 
of nucleotide frequencies compares to the plot that we obtained with the observed
*C. elegans* data. To make these plots, I first simulated 20 samples under the 
null model that the distribution is uniform across the four nucleotides, using 
the `rmultinom` function with the `size` argument set to the number of nucleotides in 
the original *C. elegans* genome data and the `prob` argument set to have an equal 
probability of each nucleotide at each spot on the genome:

```{r}
(sim_nuc_freq <- rmultinom(n = 20, 
                          size = sum(nuc_freq_df$n), 
                          prob = rep(1 / 4, 4)))
```

Next, I moved this into a tibble so I could more easily rearrange and plot the data using
facetting in `ggplot2`: 

```{r warning = FALSE, message = FALSE}
sim_nuc_freq_df <- as_tibble(sim_nuc_freq) %>% 
  mutate(nucleotide = c("A", "C", "G", "T")) %>% 
  pivot_longer(-nucleotide, names_to = "sample") %>% 
  mutate(sample = sample %>% str_remove("V") %>% as.numeric()) %>% 
  arrange(sample, nucleotide)

sim_nuc_freq_df %>% 
  slice(1:10)

ggplot(sim_nuc_freq_df, aes(x = nucleotide, y = value)) + 
  geom_col(fill = "lavender", color = "black") + 
  theme_classic() + 
  scale_y_continuous(label = scales::comma) + 
  theme(axis.title = element_blank()) + 
  labs(title = "Simulated neucleotide frequencies under a uniform model") +
  facet_wrap(~ sample) + 
  expand_limits(y = max(nuc_freq_df$n))
  
```

The y-axis limits were expanded here to cover the same range as that shown for the 
observed *C. elegans* nucleotide frequencies, to help make it easier to compare these plots
with the plot of our observed data. These plots of data simulated under the null model do 
show some variation in frequencies among the nucleotides, but it's certainly much less than 
in the observed data for *C. elegans*.

Next, I repeated this simulation process, but I increased the number of simulations to 1,000: 

```{r}
sim_nuc_freq_df <- rmultinom(n = 1000, 
                          size = sum(nuc_freq_df$n), 
                          prob = rep(1 / 4, 4)) %>% 
  as_tibble() %>% 
  mutate(nucleotide = c("A", "C", "G", "T")) %>% 
  pivot_longer(-nucleotide, names_to = "sample") %>% 
  mutate(sample = sample %>% str_remove("V") %>% as.numeric()) %>% 
  arrange(sample, nucleotide)

sim_nuc_freq_df %>% 
  slice(1:10)
```

Using this dataframe of simulations, we can measure the mean, minimum, and maximum frequencies
of each nucleotide across all 1,000 simulations:

```{r}
(sim_summary <- sim_nuc_freq_df %>% 
  group_by(nucleotide) %>% 
  summarize(mean_freq = mean(value),
            min_freq = min(value), 
            max_freq = max(value)))
```

To help compare this with the observed data, we can create a table with information from 
both the original data and the simulations under the null model: 

```{r}
nuc_freq_df %>% 
  left_join(sim_summary, by = "nucleotide") %>% 
  mutate_at(c("mean_freq", "min_freq", "max_freq", "n"), 
            prettyNum, big.mark = ",", digits = 0) %>% 
  mutate(simulations = paste0(mean_freq, " (", min_freq, ", ", max_freq, ")")) %>% 
  select(nucleotide, n, simulations) %>% 
  kable(col.names = c("Nucleotide",              
        "Frequency in C. elegans genome",
        "Mean frequency (minimum frequency, maximum frequency) across 1,000 simulations"), 
        align = "c")
```

This helps clarify how unusual the observed data would be under the null model---the 
counts of all four nucleotides in the *C. elegans* genome are completely outside the
range of frequencies in the simulated data.

Another way to look at this is with histograms of the distribution of frequencies
of each nucleotide under the null model compared to the observed frequencies in 
the *C. elegans* nucleotide: 

```{r}
ggplot(sim_nuc_freq_df, aes(x = value)) + 
  geom_histogram(binwidth = 10) + 
  facet_wrap(~ nucleotide) + 
  theme_classic() + 
  scale_x_continuous(name = "Frequency of nucleotide in the simulation under the null model",
                     labels = scales::comma) + 
  scale_y_continuous(name = "# of simulations (out of 1,000)") + 
  geom_vline(data = nuc_freq_df, aes(xintercept = n), color = "red") + 
  labs(title = expression(paste("Nucleotide frequency in ",
                                italic("C. elegans"), 
                                " compared null model simulations")),
       caption = "Red line shows the frequency observed for the nucleotide in C. elegans")
```

Finally, to help in answering this question, it would be interesting to look at a 
single measure for each simulation (and for the observed data) rather than comparing
each nucleotide one at a time. Chapter 1 gives the equation for a statistic to 
measure variability in multinomial data by calculating the sum of squares for the 
differences between the observed and expected count of nucleotides for each of the 
four nucleotides in a sample (p. 12). 

I calculated this statistic for the observed data and then for each of the 1,000
simulations.

```{r}
(obs_stat <- nuc_freq_df %>% 
  mutate(expected = mean(n),
         stat_input = (n - expected) ^ 2 / expected) %>% 
  summarize(variability_stat = sum(stat_input)))

sim_stat <- sim_nuc_freq_df %>% 
  mutate(expected = mean(value), 
         stat_input = (value - expected) ^ 2 / expected) %>% 
  group_by(sample) %>% 
  summarize(variability_stat = sum(stat_input))

sim_stat %>% 
  slice(1:5)
```

Here is a plot of the distribution of this statistic across the 1,000 simulations:

```{r}
ggplot(sim_stat, aes(x = variability_stat)) + 
  geom_rect(data = sim_stat, aes(xmin = quantile(variability_stat, prob = 0.025),
                                 xmax = quantile(variability_stat, prob = 0.975),
                                 ymin = 0, ymax = Inf), 
            fill = "beige", alpha = 0.5) +
  geom_histogram(bins = 30, fill = "white", color = "tan", alpha = 0.5) +
  theme_classic() + 
  labs(title = "Variability from expected values",
       subtitle = "Values from simulations under the null",
       x = "Value of variability statistic", 
       y = "Number of simulations with given value",
       caption = "The shaded yellow area shows the region of the central 95% of\nstatistic values for the 1,000 simulations under the null model.")

  
```
 
The value of this statistic for the observed nucleotide frequencies for *C.
elegans* is `r round(obs_stat)`, which is much larger (indicating greater variability
from expected values under the null model) than the value observed under most of
the simulations. It is, in fact, far outside the central 95% range of values
observed in simulations.
