---
title: "Vocabulary for Chapter 6"
author: "Daniel Dean"
date: '2020-03-06'
slug: vocabulary-for-chapter-6
draft: no
categories:
  - Chapter 6
  - vocabulary
tags:
  - Chapter 6
  - vocabulary
subtitle: ''
summary: ''
authors: [daniel-dean]
lastmod: 
  featured: no
image:
  caption: ''
focal_point: ''
preview_only: no
projects: []
---
  
Chapter 6 covers Statistical Testing, including a review of null and alternative hypotheses (and associated distributions), types of error (I and II), as well as challenges and opportunities introduced by multiple testing. 

```{r, echo=F, message=F, warning=F}
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)
```

```{r, echo=F, message=F, warning=F}
vocab <- read_tsv("vocab_lists/chapter_6.tsv",col_names=c("term","definition")) 

vocab %>% 
  kable(align=c("rl"), col.names=c("","")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed")) %>%
  column_spec(1, bold=T, border_right=T) %>%
  column_spec(2, width="30em")
```

### Sources Consulted or Cited

*Some of the definitons above are based in part or whole on listed definitions in the following sources:*
  
  - Holmes and Huber, 2019. *Modern Statistics for Modern Biology.* Cambridge University Press, 
Cambridge, United Kingdom.
- Wikipedia: The Free Encyclopedia. http://en.wikipedia.org/wiki/Main_Page
- Bourgon, R., Gentleman, R. & Huber, W. Independent filtering increases detection power for high-throughput experiments. Proceedings of the National Academy of Sciences 107, 9546–9551 (2010).
- Ignatiadis, N., Klaus, B., Zaugg, J. et al. Data-driven hypothesis weighting increases detection power in genome-scale multiple testing. Nat Methods 13, 577–580 (2016). https://doi.org/10.1038/nmeth.3885
- https://www.statisticssolutions.com/bonferroni-correction/
- https://bioconductor.org/packages/release/bioc/vignettes/IHW/inst/doc/introduction_to_ihw.html
- https://www.statisticshowto.datasciencecentral.com/familywise-error-rate/
- https://www.statisticshowto.datasciencecentral.com/probability-and-statistics/hypothesis-testing/
- https://www.biostars.org/p/273537/

  

  
  ### Practice
  
  <iframe src="https://quizlet.com/488864042/flashcards/embed?i=2oqpc3&x=1jj1" height="500" width="100%" style="border:0"></iframe>
  