---
title: Vocabulary for Chapter 12
author: Sherry WeMott
date: '2020-05-04'
slug: vocabulary-for-chapter-12
categories:
  - vocabulary
  - Chapter 12
tags:
  - vocabulary
subtitle: ''
summary: ''
authors: []
lastmod: '2020-05-04T19:46:32-06:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
Chapter 12 covers supervised learning and the statistics of predicting categorical variables. Also discussed are the issues of overfitting and generalizability and how to "train" statistical models. 

```{r, echo=F, message=F, warning=F}
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)
```

```{r, echo=F, message=F, warning=F}
vocab <- read_tsv("vocab_lists/chapter_12.tsv",col_names=c("term","definition")) 

vocab %>% 
    kable(align=c("rl"), col.names=c("","")) %>%
    kable_styling(bootstrap_options = c("striped","hover","condensed")) %>%
    column_spec(1, bold=T, border_right=T) %>%
    column_spec(2, width="30em")
```
The vocabulary words for Chapter 12 are: 

### Sources consulted or cited 

*Some of the definitons above are based in part or whole on listed definitions in the following source:*

- Holmes and Huber, 2019. *Modern Statistics for Modern Biology.* Cambridge University Press, 
Cambridge, United Kingdom.

https://en.wikipedia.org/
https://machinelearningmastery.com/overfitting-and-underfitting-with-machine-learning-algorithms/
https://.statisticshowto.com
https://www.cs.cmu.edu/~schneide/tut5/node42.html
https://towardsdatascience.com
https://bioconductor.org
https://pfern.gtihub.io

### Practice

<iframe src="https://quizlet.com/506392624/match/embed?i=takib&x=1jj1" height="500" width="100%" style="border:0"></iframe>

