---
title: Vocabulary for Chapter 1
author: Brooke Anderson
date: '2020-01-13'
slug: vocabulary-for-chapter-1
categories:
  - vocabulary
  - Chapter 1
tags:
  - vocabulary
  - Chapter 1
subtitle: ''
summary: ''
authors: []
lastmod: '2020-01-13T16:20:46-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

Chapter 1 covers generative modeling for discrete data. It introduces a number of terms covering
probablity and statistical modeling, as well as a few biological terms. The vocabulary words for 
Chapter 1 are: 

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(readr)
vocab <- read_tsv("vocab_lists/chapter_1.tsv", col_names = c("term", "definition"))
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(kableExtra)
vocab %>% 
  kable(align = c("rl"), 
        col.names = c("", "")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover",
                                      "condensed")) %>% 
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2, width = "30em")
```

### Sources consulted or cited 

*Some of the definitions above are based in part or whole on listed definitions 
in the following sources.*

- Holmes and Huber, 2019. *Modern Statistics for Modern Biology.* Cambridge University Press, 
Cambridge, United Kingdom.
- Everitt and Skrondal, 2010. *The Cambridge Dictionary of Statistics (Fourth Edition).* Cambridge University Press, Cambridge, United Kingdom.
- Bioconductor: Open Source Software for Bioinformatics. https://www.bioconductor.org/
- Wikipedia: The Free Encyclopedia. https://en.wikipedia.org/wiki/Main_Page
- NIH Genetics Home Reference. https://ghr.nlm.nih.gov/
- NCI Dictionary of Cancer Terms. https://www.cancer.gov/publications/dictionaries/cancer-terms 

### Practice

<iframe src="https://quizlet.com/473473135/flashcards/embed?i=2ldef3&x=1jj1" height="500" width="100%" style="border:0"></iframe>

