---
title: "Vocabulary for Chapter 7"
author: "Zach Laubach"
date: '2020-03-30'
slug: vocabulary-for-chapter-7
draft: no
categories:
  - Chapter 7
  - vocabulary
tags:
  - Chapter 7
  - vocabulary
subtitle: ''
summary: ''
authors: [zach_laubach]
lastmod: '2020-03-30T21:37:37-06:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

Chapter 7 covers multivariate analysis, with a focus on principal component analysis and
dimension reduction in general.

```{r, echo=F, message=F, warning=F}
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)
```

```{r, echo=F, message=F, warning=F}
vocab <- read_tsv("vocab_lists/chapter_7.tsv",col_names=c("term","definition")) 

vocab %>% 
    kable(align=c("rl"), col.names=c("","")) %>%
    kable_styling(bootstrap_options = c("striped","hover","condensed")) %>%
    column_spec(1, bold=T, border_right=T) %>%
    column_spec(2, width="30em")
```

### Source Consulted or Cited

*Some of the definitons above are based in part or whole on listed definitions in the following source:*

- Holmes and Huber, 2019. *Modern Statistics for Modern Biology.* Cambridge University Press, 
Cambridge, United Kingdom.
- Wikipedia: The Free Encyclopedia. http://en.wikipedia.org/wiki/Main_Page

### Practice

<iframe src="https://quizlet.com/499545557/flashcards/embed?i=2ldef3&x=1jj1" height="500" width="100%" style="border:0"></iframe>
