---
title: Vocabulary for Chapter 1
author: Brooke Anderson
date: '2020-01-13'
slug: vocabulary-for-chapter-1
categories:
  - vocabulary
  - Chapter 1
tags:
  - vocabulary
  - Chapter 1
subtitle: ''
summary: ''
authors: []
lastmod: '2020-01-13T16:20:46-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
vocab <- tribble(
  ~term, ~definition,
  
  "probability model", "",
  
  "vector", "",
  
  "discrete event", "",
  
  "categorical variable", "A variable that can belong to one of a finite set of levels.",
  
  "levels", "In the context of a categorical variable, the set of groups to which the variable can be assigned.",
  
  "factor", "",
  
  "exchangeable", "A way to describe a random variable for which the order that the data are observed doesn't matter",
  
  "sufficient statistic", "A statistic that, even though it is a summary, contains all the information that you would use from an original, uncondensed form of the data. An example is frequency counts for an exchangeable ...",
  
  "Bernoulli distribution", "A probability distribution to describe the outcome probability for two possible events.",
  
  "parameter", "A numerical value that describes a population or a model.",
  
  "complementary", "A description of two events who are mutually exclusive and whose probabilities sum to one. In other words, either one event or the other is guaranteed to happen, but not both.",
  
  "binomial random variable", "A variable whose values occur according to a binomial probability distribution.",
  
  "probability mass distribution", "",
  
  "Poisson distribution", "A probability distribution that is appropriate when the probability of success is small and the number of trials is large.",
  
  "epitope", "Site on a macromolecular antigen to which an antibody binds. This is the part of an antigen that is recognized by the immune system and is also known as an antigen determinent.",
  
  "Enzyme-linked immunosorbent assay (ELISA)", "An assay that is used to detect specific epitopes at different positions along a protein.", 
  
  "conditional on", "Given",
  
  "cumulative distribution function", "",
  
  "extreme value analysis", "Analysis focused on the behavior of the very large or the very small values of a random distribution, allowing an exploration of the probability of rare events.", 
  
  "rare events", "Something that occurs with a very low probability.",
  
  "rank statistic", "A statistic based on the order of the relative positions of observations based on a certain characteristics.", 
  
  "Monte Carlo method", "A method that uses a computer simulation based on our generative model that finds the probabilities of the events we're interested in.",
  
  "probability or generative modeling", "A method of modeling where all the parameters are known and the mathematical theory allows us to work by deduction.", 
  
  "deduction", "A top-down method of reasoning ...",
  
    "statistical modeling", "A method of modeling where the distribution of the data is not known.",
  
  "fit", "In the context of data and a statistical distribution, ...",
  
  "multinomial", "",
  
  "power", "The probability of detecting something if it is there.",
  
  "true positive rate", "The probability of detecting something if it is there.", 
  
  "null hypothesis", "", 
  
  "matrix", "", 
  
  "expected value", "",
  
  "variability", "", 
  
  "statistic", "A numerical characteristic of a sample",
  
  "null distribution", "", 
  
  "alternative", "In the context of a generating process, ...", 
  
  "chi-squared distribution", "", 
  
  "p-value", "The probability of seeing the data that were observed given a hypothesis.", 
  
  "probability density function", "",
  
  "default", "In the context of function parameters for an R function, the value that is used if no custom value is specified.", 
  
  "C. elegans genome nucleotide frequency", "",
  
  "Bioconductor", ""
  
)
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(kableExtra)
vocab %>% 
  kable(align = c("rl"), 
        col.names = c("", "")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover",
                                      "condensed")) %>% 
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2, width = "30em")
```

### Sources consulted or cited 

*Some of the definitions above are based in part or whole on listed definitions 
in the following sources.*

- Holmes and Huber, 2019. *Modern Statistics for Modern Biology.* Cambridge University Press, 
Cambridge, United Kingdom.
- Everitt and Skrondal, 2010. *The Cambridge Dictionary of Statistics (Fourth Edition).* Cambridge University Press, Cambridge, United Kingdom.

### Practice

<iframe src="https://quizlet.com/473473135/flashcards/embed?i=2ldef3&x=1jj1" height="500" width="100%" style="border:0"></iframe>

