---
title: "Vocabulary for Chapter 4"
author: "Amy Fox"
date: '2020-02-20'
slug: vocabulary-for-chapter-4
categories:
  - vocabulary
  - Chapter 4
tags:
  - vocabulary
  - Chapter 4
subtitle: ''
summary: ''
authors: []
lastmod: 
  featured: no
image:
  caption: ''
focal_point: ''
preview_only: no
projects: []
---
  
Chapter 4 covers how to generate both finite and infinite mixture models from various distributions. It introduces a number of terms relating to these models. The vocabulary words for Chapter 4 are: 
  
```{r echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(readr)
vocab <- read_tsv("vocab_lists/chapter_4.tsv", col_names = c("term", "definition"))
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(kableExtra)
vocab %>% 
  kable(align = c("rl"), 
        col.names = c("", "")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover",
                                      "condensed")) %>% 
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2, width = "30em")
```

### Sources consulted or cited 

*Some of the definitions above are based in part or whole on listed definitions 
in the following sources.*
  
  - Holmes and Huber, 2019. *Modern Statistics for Modern Biology.* Cambridge University Press, 
Cambridge, United Kingdom.
- Everitt and Skrondal, 2010. *The Cambridge Dictionary of Statistics (Fourth Edition).* Cambridge University Press, Cambridge, United Kingdom.
- *Zero-Inflated Poisson Regression.* Institute for Digital Research and Education Statistical Consulting. https://stats.idre.ucla.edu/r/dae/zip/.
- Berrar, 2019. *Introduction to Non-parametric Bootstrap.* Research Gate. https://www.researchgate.net/
- Do and Batzoglou, 2008. *What is the expectaion maximization algorithm?.* Nature Biotechnology. 
- Wikipedia: The Free Encylcopedia. https://en.wikipedia.org/wiki/Main_Page
- Google Oxford American Dictionary. https://www.google.com
- d'Auzay, et al., 2019. *Statistics of progress variable and mixture fraction gradients in an open turbulent jet spray flame.* Fuel.
- Brownlee, 2019. *A Gentle Introduction to Expectation-Maximization (EM Algorithm).* Machine Learning Mastery. https://www.machinelearningmastery.com 
- *Non-parametric Methods.* R tutorial. https://www.r-tutor.com
- *Precise analysis of DNA–protein binding sequences.* Illumina. https://www.illumina.com
- *Microarray.* Nature. https://www.nature.com


### Practice

<iframe src="https://quizlet.com/485299672/match/embed?i=2r4ib&x=1jj1" height="500" width="100%" style="border:0"></iframe>