---
title: Vocabulary for Chapter 5
author: "Burton Karger"
date: '2020-02-21'
slug: vocabulary-for-chapter-5
categories:
  - Chapter 5
  - vocabulary
tags:
  - Chapter 5
  - vocabulary
subtitle: ''
summary: ''
authors: [burton-karger]
lastmod: '2020-02-21T14:41:22-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
Chapter 5 covers Clustering Analysis for large scale data anlysis like DNA/RNA sequencing outputs. These methods produce so much data that more unbiased approaches are required when attempting to make correlations. 

```{r, echo=F, message=F, warning=F}
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)
```

```{r, echo=F, message=F, warning=F}
vocab <- read_tsv("vocab_lists/chapter_5.tsv",col_names=c("term","definition")) 

vocab %>% 
    kable(align=c("rl"), col.names=c("","")) %>%
    kable_styling(bootstrap_options = c("striped","hover","condensed")) %>%
    column_spec(1, bold=T, border_right=T) %>%
    column_spec(2, width="30em")
```

### Sources Consulted or Cited

*Some of the definitons above are based in part or whole on listed definitions in the following sources:*

- Holmes and Huber, 2019. *Modern Statistics for Modern Biology.* Cambridge University Press, 
Cambridge, United Kingdom.
- Wikipedia: The Free Encyclopedia. http://en.wikipedia.org/wiki/Main_Page
- https://academic.oup.com/biolreprod/article/85/5/946/2530522
- https://discovery.lifemapsc.com/in-vivo-development/inner-cell-mass/inner-cell-mass
- https://study.com/academy/lesson/inner-cell-mass-icm-definition-function-quiz.html
- https://www.sciencedirect.com/
- https://www.medicinenet.com/
- https://www.niaid.nih.gov/
- https://iti.stanford.edu/
- https://sysbiowiki.soe.ucsc.edu/node/323
- https://www.statisticshowto.datasciencecentral.com/between-group-variation/
- https://vsoch.github.io/2013/the-gap-statistic/

### Practice

<iframe src="https://quizlet.com/488864042/match/embed?i=2oqpc3&x=1jj1" height="500" width="100%" style="border:0"></iframe>
